\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2020-03-19 sections/abstract.tex
Updated draft of manuscript
Edits made by Daniel [03-16-2020]

\begin{abstract}
  {
    \noindent
    Microbes tend to organize into communities consisting of hundreds of species involved in complex interactions with each other.
    16S ribosomal RNA (16S rRNA) gene profiling provides snapshots that reveal the phylogenies and abundance profiles of these microbial communities.
    These snapshots, when collected from multiple samples, have the potential to reveal which microbes co-occur, providing a glimpse into the network of influences in these communities.
    The inference of networks from 16S data is prone to statistical artifacts.
    There are many tools for performing each step of the 16S analysis workflow and the results generated by different tools are varied, but the extent to which these steps affect the resultant network is still unclear.
    In this study, we perform a meticulous analysis of each step of a pipeline that processes 16S sequencing data into a network of microbial associations.
    Through this process, we determine the tools and parameters that generate the most accurate and robust co-occurrence networks.
    Ultimately, we develop a standardized pipeline that follows these default tools and parameters, but that can also help explore the outcome of any other combination of choices.
    We envisage that this standard pipeline for processing 16S sequencing data into networks of microbial co-occurrences could be used for integrating multiple data-sets, and for generating comparative analyses and consensus networks useful for detecting disease-related patterns. \\
    The pipeline is available as a Python package at \href{https://github.com/dileep-kishore/mindpipe}{https://github.com/dileep-kishore/mindpipe}
  }
\end{abstract}

>>>>

\begin{abstract}
  {
    \noindent
    Microbes tend to organize into communities consisting of hundreds of species involved in complex interactions with each other.
    16S ribosomal RNA (16S rRNA) amplicon profiling provides snapshots that reveal the phylogenies and abundance profiles of these microbial communities.
    These snapshots, when collected from multiple samples, have the potential to reveal which microbes co-occur, providing a glimpse into the network of influences in these communities.
    The inference of networks from 16S data is prone to statistical artifacts.
    There are many tools for performing each step of the 16S analysis workflow, but the extent to which these steps affect the final network is still unclear.
    In this study, we perform a meticulous analysis of each step of a pipeline that processes 16S sequencing data into a network of microbial associations.
    % TODO: Instead of this say: We analyze the parts of the process that cause the most variance and what is the result
    Through this process, we map how different choices of algorithms and parameters affect the co-occurrence network, and estimate steps which contribute most significantly to the variance. We further determine the tools and parameters that generate the most accurate and robust co-occurrence networks, based on comparison with mock and synthetic datasets.
    Ultimately, we develop a standardized pipeline that follows these default tools and parameters, but that can also help explore the outcome of any other combination of choices.
    % TODO: Make this general (do not confine it to disease-related) - niche-related(?)
    We envisage that this standard pipeline for processing 16S sequencing data into networks of microbial co-occurrences could be used for integrating multiple data-sets, and for generating comparative analyses and consensus networks useful for detecting disease-related patterns. \\
    % TODO: Move package to segrelab github
    The pipeline is available as a Python package at \href{https://github.com/dileep-kishore/mindpipe}{https://github.com/dileep-kishore/mindpipe}
  }
\end{abstract}

### 2020-03-19 sections/introduction.tex
Updated draft of manuscript
Edits made by Daniel [03-16-2020]

  High-throughput metagenomic sequencing techniques (such as 16S rRNA gene profiling) help detect, identify and quantify a large part of the constitutive microorganisms of a microbiome \cite{Jovel2016}.
  Knowledge of the community composition at a particular instance in time would enable us to derive partial insights into the underlying dynamics.
  With the advancement in DNA sequencing technologies \cite{Narihiro2017} and data processing methods much more information can be extracted from these microbial community samples than ever before.
  These advances have led to large-scale data collection efforts involving environmental (Earth Microbiome Project) \cite{Thompson2017}, marine (Tara Oceans Project) \cite{Zhang2015} and human-associated microbiota (Human Microbiome Project) \cite{HumanMicrobiomeProjectConsortium2012} and thus led the generation of a wealth of data about these microbial communities.
  These data-sets can be used to infer networks of co-occurrence relationships between microorganisms in the samples.
  The co-occurrence networks can in turn be used to identify important interactions driving the dynamics of the microbial community, as well as, analyze emergent properties such as robustness and modularity.

  16S sequencing data is limited by many factors such as resolution, sequencing depth, compositional nature, sequencing errors and copy number variations.
  Many different analysis methods have been developed \cite{Callahan2016,Amir2017,Friedman2012,Kurtz2015} to counter these limitations and remove statistical artifacts, leading to the existence of a myriad of tools for each step of the data analysis workflow.
  The different methods and tools developed to solve these issues infer vastly different community compositions and co-occurrence networks \cite{Golob2017,Weiss2016}, making the analysis unreproducible and comparisons convoluted.
  But because there does not exist a dataset wherein the microbial interactions have been fully characterized, there is no straightforward way to determine which set of tools or methods generate the most accurate networks.
  In this study, we propose to build a standardized 16S data-analysis pipeline that will produce robust and reproducible co-occurrence networks from community 16S sequence data.
  We do this by conducting a systematic study of the various methods currently available.
  We compare different methods to each other and additionally use mock microbial community datasets and synthetically generated reads to benchmark them.

  There exist many platforms like MG-RAST \cite{Keegan2016}, Qiita \cite{qiita} and tools such as QIIME \cite{Caporaso2010} that provide pipelines for 16S data analysis, but the main motivations behind the construction of our own pipeline are two-fold.
  Firstly, the existing tools do not focus their analysis on co-occurrence networks and the effect of upstream statistical methods on the inferred networks.
  Secondly, we undertake a comprehensive study of existing methods in order to suggest optimal methods, tools and parameters for users of the pipeline.
  We believe that these steps will ensure better reproducibility and easier comparison of networks across data-sets.
  We call our pipeline mindpipe, which derives its origin from the \ac{mind}.

>>>>

  High-throughput metagenomic sequencing techniques (such as 16S rRNA gene profiling) help detect, identify and quantify a large part of the constitutive microorganisms of a microbiome \cite{Jovel2016}.
  Knowledge of the community composition at a particular instance in time would enable us to derive partial insights into the underlying dynamics.
  With the advancement in DNA sequencing technologies \cite{Narihiro2017} and data processing methods much more information can be extracted from these microbial community samples than ever before.
  These advances have led to large-scale data collection efforts involving environmental (Earth Microbiome Project) \cite{Thompson2017}, marine (Tara Oceans Project) \cite{Zhang2015} and human-associated microbiota (Human Microbiome Project) \cite{HumanMicrobiomeProjectConsortium2012} and thus led the generation of a wealth of data about these microbial communities.

  These data-sets can be used to infer networks of co-occurrence relationships between microorganisms in the samples.These networks have microbial taxa as nodes, and edges that represent the frequent co-occurrence (or negative correlations) across different datasets. Co-occurrence networks can in principle be used to identify important interactions driving the dynamics of the microbial community, including emergent ecosystem-level properties such as ecosystem robustness and modularity.

  One of the limitations of this type of analysis is that 16S sequencing data is limited by many factors such as resolution, sequencing depth, compositional nature, sequencing errors and copy number variations.

  Many different analysis methods have been developed \cite{Callahan2016,Amir2017,Friedman2012,Kurtz2015} to counter these limitations and remove statistical artifacts, leading to the existence of a myriad of tools for each step of the data analysis workflow.
  The different methods and tools developed to solve these issues infer vastly different community compositions and co-occurrence networks \cite{Golob2017,Weiss2016}, making it difficult to reliably compare networks across different publications and studies. Conversely, given the lack of comprehensive comparisons between directly observed microbial interactions (e.g. from co-culture experiments) and co-occurence networks, there is no straightforward way to determine which set of tools or methods generate the most accurate networks.

 Previous work has developed popular platforms (like MG-RAST \cite{Keegan2016}, Qiita \cite{qiita}) and tools (such as QIIME \cite{Caporaso2010}) that provide pipelines for 16S data analysis. However, existing tools typically are not focused on the effects of upstream statistical methods on the inferred co-occurrence networks. Furthermore, no organized framework currently exist to systematically analyze and compare existing components of the data-analysis from amplicons to networks.

  In this study, we present a standardized 16S data-analysis pipeline that produces robust and reproducible co-occurrence networks from community 16S sequence data, and allow users to interactively explore how the network would change upon using different alternative tools and parameters at each step. Our pipeline is coupled to an online integrative tool for the organization, visualization and analysis of inter-microbial networks. Given that this 
  In addition to making this tool freely available, we implemented a systematic comparative analysis to determine which steps of the pipeline have the largest influence on the final network, and what choice seems to provide best agreement with the tested mock and synthetic datasets.
  We believe that these steps will ensure better reproducibility and easier comparison of co-occurrence networks across data-sets. We expect that our tool will also be useful for benchmarking future alternative methods, and for ensuring a transparent evaluation of the possible biases introduced by the use of specific tools.

  % TODO: Study the diversity of methods and explore the possiblity of a standardized approach to 16S analysis
  In this study, we propose to build a standardized 16S data-analysis pipeline that will produce robust and reproducible co-occurrence networks from community 16S sequence data.
  We do this by conducting a systematic study of the various methods currently available.
  We compare different methods to each other and additionally use mock microbial community datasets and synthetically generated reads to benchmark them.

  Secondly, we undertake a comprehensive study of existing methods in order to suggest optimal methods, tools and parameters for users of the pipeline.
  We believe that these steps will ensure better reproducibility and easier comparison of networks across data-sets.
  We call our pipeline mindpipe, which derives its origin from the \ac{mind}.

### 2020-03-19 sections/methods.tex
Update methods and results
Edits made by Daniel [03-16-2020]

\section*{Materials and Methods}

@@

  \vspace{-5mm}
  The study uses three kinds of 16S rRNA sequencing datasets \- real datasets, mock datasets and synthetic datasets.
  Real datasets are those sequencing reads obtained from naturally occurring microbial community samples.
  The current study used healthy stool samples from a fecal microbiome transplant study \cite{Kang2017} and healhty saliva samples from a periodontal disease study~\cite{Chen2018} as real datasets for analysis.
  The mock community 16S datasets were obtained from mockrobiota~\cite{Bokulich2016}.
  The mock datasets used for this study were labelled mock4, mock12 and mock16.
  The mock4 community is composed of 21 bacterial strains represented in equal abundances in two replicate samples, and same strains in unequal abundances in two more replicate samples.
  The mock12 community is composed of 27 bacterial strains that include closely related taxa with some pairs having only one to two nucleotide difference from another.
  The mock16 community is composed of equal amounts of 49 bacteria and 10 archea.
  The synthetic datasets were generated using an artificial read simulator called ART~\cite{Huang2012}.
  Three different microbial composition profiles were used as input
  Synthetic reads were generated using a soil and water microbiome composition profiles from the Earth Microbiome Project~\cite{Thompson2017} and healthy gut microbiome project from the Human Microbiome Project~\cite{HumanMicrobiomeProjectConsortium2012}.
  The scripts used to generate the synthetic data are in the scripts folder of the respository (\href{https://github.com/dileep-kishore/mind-pipeline-paper}{https://github.com/dileep-kishore/mind-pipeline-paper})

>>>>

% TODO: Make the methods more mathematically inclined
\section*{Materials and Methods}

@@

  \vspace{-5mm}
  The study uses three kinds of 16S rRNA sequencing datasets \- real datasets, mock datasets and synthetic datasets.
  Real datasets are those sequencing reads obtained from naturally occurring microbial community samples.
  The current study used healthy stool samples from a fecal microbiome transplant study \cite{Kang2017} and healthy saliva samples from a periodontal disease study~\cite{Chen2018} as real datasets for analysis.
  The mock community 16S datasets were obtained from mockrobiota~\cite{Bokulich2016}.
  The mock datasets used for this study were labelled mock4, mock12 and mock16.
  The mock4 community is composed of 21 bacterial strains represented in equal abundances in two replicate samples, and same strains in unequal abundances in two more replicate samples.
  The mock12 community is composed of 27 bacterial strains that include closely related taxa with some pairs having only one to two nucleotide difference from another.
  The mock16 community is composed of equal amounts of 49 bacteria and 10 archea.
  The synthetic datasets were generated using an artificial read simulator called ART~\cite{Huang2012}.
  Three different microbial composition profiles were used as input
  Synthetic reads were generated using a soil and water microbiome composition profiles from the Earth Microbiome Project~\cite{Thompson2017} and healthy gut microbiome project from the Human Microbiome Project~\cite{HumanMicrobiomeProjectConsortium2012}.
  The scripts used to generate the synthetic data are in the scripts folder of the respository (\href{https://github.com/dileep-kishore/mind-pipeline-paper}{https://github.com/dileep-kishore/mind-pipeline-paper})
